Visual proteomics using whole-lamella 2D template matching

### 2022-03-07 manuscript.tex
Nikos comments [ci skip]
This build is based on
https://github.com/jojoelfe/deco_lace_template_matching_manuscript/commit/287b9950fc2bff7878d36aa1cbf2cb7f2082f798.

This commit was created by the following CI build and job:
https://github.com/jojoelfe/deco_lace_template_matching_manuscript/commit/287b9950fc2bff7878d36aa1cbf2cb7f2082f798/checks
https://github.com/jojoelfe/deco_lace_template_matching_manuscript/actions/runs/1948085649

This manuscript
(\href{https://jojoelfe.github.io/deco_lace_template_matching_manuscript/v/a4570afcb2725b9980fa97358516050b0a2697d3/}{permalink})
was automatically generated
from \href{https://github.com/jojoelfe/deco_lace_template_matching_manuscript/tree/a4570afcb2725b9980fa97358516050b0a2697d3}{jojoelfe/deco\_lace\_template\_matching\_manuscript@a4570af}
on March 3, 2022.

@@

Localization of biomolecules inside a cell is an important goal of biological imaging. Fluorescence microscopy can localize biomolecules inside whole cells and tissues, but its ability to count biomolecules and accuracy of the spatial coordinates is limited by the wavelength of visible light. On the other hand, cryo-electron microscopy (cryo-EM) provides highly accurate position and orientation information of biomolecules but is often limited to small fields of view inside a cell, providing only limited biological context. In this study we use a new data-acquisition scheme called ``Beam-Imageshift for Large Area Cryo-Electron microscopy'' (BILACE) to collect high-resolution cryo-EM data over entire sections (100 -- 200 nm thick lamellae) of neutrophil-like mouse cells, representing roughly 1\% of the total cellular volume. We use 2D template matching to determine localization and orientation of the large ribosomal subunit in these sections. Furthermore, we use 2D template matching to test detected targets for the presence of the small ribosomal subunit and used the relative orientations of the ribosomes to assign them to polysomes. These results provide ``maps'' of translational activity across sections of mammalian cells. We envision that using this high-throughput cryo-EM data collection approach together with 2D template matching will advance visual proteomics to complement other single-cell ``omics'' techniques, such as flow-cytometry and single-cell sequencing.

@@

Understanding of cellular processes requires knowledge of the amounts, location,
interactions, and conformations of biomolecules inside the cell. Techniques that
measure this can broadly be divided into label- and label-free techniques. In
label-techniques a probe is physically attached to a molecule of interest that
is able to produce a high signal-to-noise signal, such as a fluorescent molecule.
In label-free techniques the physical properties of molecules themselves are
used for detection. An example for this is proteomics using mass-spectrometry
{[}\protect\hyperlink{ref-tSXIKPl7}{1}{]}. The advantage of label-free techniques is that
they can provide information over thousands of molecules, while label-techniques
offer higher fidelity information for a few molecules. Especially spatial
information can most of the time only be achieved using label-techniques, such
as fluorescence microscopy {[}\protect\hyperlink{ref-VBmW7Aot}{2}{]}.

Cryo-electron microscopy has the potential to directly measure the arrangement
of atoms that compose biomolecules inside of cells, thereby allowing label-free
detection with high spatial accuracy. This has been called ``visual proteomics''
{[}\protect\hyperlink{ref-tGQ6TSUo}{3}{]}. Since cryo-EM requires thin samples (\textless500nm), imaging of
biomolecules inside cells is either restricted to small organisms, thin
regions of large cells, or requires thinning of the sample. This can be achieved
either by mechanical sectioning {[}\protect\hyperlink{ref-g8QavfwP}{4}{]} or by
milling using a focused ion beam (FIB) {[}\protect\hyperlink{ref-16IhS1Nc4}{5}{]}. This
complex workflow restricts throughput of cryo-EM imaging of cells. This is
exacerbated by the fact that at the required magnifications, typical field of
views (FOV) are very small compared to mammalian cells and the FOV achieved by
label-techniques such as fluorescence light microscopy. The predominant cryo-EM
technique for detection of biomolecules according to their shape in cells at the
moment is cryo-electron tomography {[}\protect\hyperlink{ref-Rksh2dxu}{6}{]}. However,
the requirement of physically tilt the stage at every FOV, together with a more
complex workflow that requires highly accurate alignment of various projection,
further restricts the throughput for molecular detection.

An alternative approach is to identify molecules by their structural
``fingerprint'' in single projection using ``2D template-matching''
{[}\protect\hyperlink{ref-Ynb3IP6I}{7},\protect\hyperlink{ref-18KGpXYPE}{8},\protect\hyperlink{ref-10bXZuF3G}{9}{]}. In this method an experimentally obtained 3D model
of a biomolecule is used to calculate the expected electron density, which is
called the template. The template is then projected on a fine angular grid and
the projections are used to find local cross-correlation peaks in a cryo-EM
micrograph. Since locations of the biomolecule in the Z-direction causes
predictable aberrations to the projection image, this method can be used to
calculate 3D coordinates and orientations of a biomolecule in a cellular sample
{[}\protect\hyperlink{ref-18KGpXYPE}{8}{]}

@@

Here we apply 2D-template matching of ribosomes to cryo-FIB milled
neutrophil-like murine cells {[}\protect\hyperlink{ref-1B9Vt9eYu}{12}{]}. To increase the
amount of collected data and to provide un-biased sampling of the whole lamella
we devised a new data-acquisition scheme, Defocus-corrected large area
cryo-electron microscopy (DeCo-LACE). We characterize aberration cause by the
used large beam-image shifts and highly focused beams and find that they can be
adequately correct to enable ribosome detection by 2D-template matching. The
resulting data provides a description of ribosome distribution in the whole
lamellae, which represent roughly 2\% of the cellular volume. We find highly
heterogeneous density of ribosome within the cell and can identify discrete
clusters of presumably translationaly active ribosomes, by testing for the
presence of the small ribosomal subunit. The high accuracy of location and
orientation of each detected ribosome also allows us to cluster ribosome
molecules into potential polysomes. Analysis of the throughput in this method
suggests that for the foreseeable future computation will be the bottleneck for
visual proteomics.

>>>>

This manuscript
(\href{https://jojoelfe.github.io/deco_lace_template_matching_manuscript/v/287b9950fc2bff7878d36aa1cbf2cb7f2082f798/}{permalink})
was automatically generated
from \href{https://github.com/jojoelfe/deco_lace_template_matching_manuscript/tree/287b9950fc2bff7878d36aa1cbf2cb7f2082f798}{jojoelfe/deco\_lace\_template\_matching\_manuscript@287b995}
on March 7, 2022.

@@

Localization of biomolecules inside a cell is an important goal of biological
imaging. Fluorescence microscopy can localize biomolecules inside whole cells
and tissues, but its ability to count biomolecules and accuracy of the spatial
coordinates is limited by the wavelength of visible light. Cryo-electron
microscopy (cryo-EM) provides highly accurate position and orientation
information of biomolecules but is often confined to small fields of view inside
a cell, limiting biological context. In this study we use a new data-acquisition
scheme called ``Defocus-Corrected Large-Area cryo-EM'' (DeCo-LACE) to collect
high-resolution cryo-EM data over entire sections (100 -- 200 nm thick lamellae)
of neutrophil-like mouse cells, representing roughly 1\% of the total cellular
volume. We use 2D template matching (2DTM) to determine localization and orientation
of the large ribosomal subunit in these sections, detect bound small
ribosomal subunits and assign ribosomes to polysomes based on their relative orientations to each other. These data provide ``maps'' of translational activity across
sections of mammalian cells. This new high-throughput cryo-EM
data collection approach together with 2DTM will advance visual
proteomics and complement other single-cell ``omics'' techniques, such as
flow-cytometry and single-cell sequencing.

@@

A major goal in understanding cellular processes is the knowledge of the
amounts, location, interactions, and conformations of biomolecules
inside the cell. This knowledge can be obtained by approaches broadly
divided into label- and label-free techniques. In label-dependent
techniques a probe is physically attached to a molecule of interest that
is able to detected with a high signal-to-noise signal, such as a
fluorescent molecule. In label-free techniques the physical properties
of molecules themselves are used for detection. An example for this is
proteomics using mass-spectrometry {[}\protect\hyperlink{ref-tSXIKPl7}{1}{]}. The advantage
of label-free techniques is that they can provide information over
thousands of molecules, while label-techniques offer highly specific
information for a few molecules. Especially spatial information can
often only be achieved using label-dependent techniques, such as
fluorescence microscopy {[}\protect\hyperlink{ref-VBmW7Aot}{2}{]}.

Cryo-electron microscopy has the potential to directly visualize the
arrangement of atoms that compose biomolecules inside of cells, thereby
allowing label-free detection with high spatial accuracy. This has been
called ``visual proteomics'' {[}\protect\hyperlink{ref-tGQ6TSUo}{3}{]}. Since cryo-EM
requires thin samples (\textless500nm), imaging of biomolecules inside cells is
restricted to small organisms, thin regions of cells, or samples that
have been suitably thinned. Thinning can be achieved either by
mechanical sectioning {[}\protect\hyperlink{ref-g8QavfwP}{4}{]} or by
milling using a focused ion beam (FIB) {[}\protect\hyperlink{ref-16IhS1Nc4}{5}{]}. his complex workflow
leads to a low throughput of cryo-EM imaging of cells and is further
limited by the fact that at the required magnifications, typical fields
of view (FOV) are very small compared to mammalian cells, and the FOV
achieved by label-techniques such as fluorescence light microscopy. The
predominant cryo-EM technique for the localization of biomolecules of
defined size and shape inside cells is cryo-electron tomography {[}\protect\hyperlink{ref-Rksh2dxu}{6}{]}. However, the requirement of a tilt series at
every imaged location and subsequent image alignment, severely limits
the throughput for molecular localization.

An alternative approach is to identify molecules by their structural
``fingerprint'' in single projection using ``2D template-matching'' (2DTM)
{[}\protect\hyperlink{ref-Ynb3IP6I}{7},\protect\hyperlink{ref-18KGpXYPE}{8},\protect\hyperlink{ref-10bXZuF3G}{9}{]}. In this
method a 3D model of a biomolecule is used as a template to find 2D
projections that match the molecules visible in the electron
micrographs. This method requires a projection search on a fine angular
grid, and the projections are used to find local cross-correlation peaks
with the micrograph. Since the location of a biomolecule in the
z-direction causes predictable aberrations to the projection image, this
method can be used to calculate complete 3D coordinates and orientations
of a biomolecule in a cellular sample
{[}\protect\hyperlink{ref-18KGpXYPE}{8}{]}

@@

Here we apply 2D-template matching of ribosomes to cryo-FIB milled
neutrophil-like murine cells {[}\protect\hyperlink{ref-1B9Vt9eYu}{12}{]}. To increase the
amount of collected data and to provide unbiased sampling of the whole
lamella, we devised a new data-acquisition scheme, Defocus-corrected
large area cryo-electron microscopy (DeCo-LACE). We characterize
aberration cause by the used large beam-image shifts and highly focused
beams and find that they can be adequately corrected to enable ribosome
detection by 2DTM. The resulting data provide a description of ribosome
distribution in the whole lamellae, which represent roughly 2\% of the
cellular volume. We find highly heterogeneous density of ribosome within
the cell and can identify discrete clusters of presumably
translationally active ribosomes, by testing for the presence of the
small ribosomal subunit. The high accuracy of location and orientation
of each detected ribosome also allows us to cluster ribosome molecules
into potential polysomes. Analysis of the throughput in this method
suggests that for the foreseeable future computation will be the
bottleneck for visual proteomics.
